Methods for black hole merger parameter estimation with LISA.

### 2020-10-13 docs/paper_LDC/paper_LDC.tex
These were uncommittted changes from last December (LDC process).  Most are the LDC paper start.  There are also some comments added in data_likelihood.c and a flag for extra checking there turned off.

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
\newcommand{\hatk}{\hat{k}}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand{\bsub}{\begin{subequations}}
\newcommand{\esub}{\end{subequations}}
\newcommand\betaL{{\beta_{L}}}
\newcommand\lambdaL{{\lambda_{L}}}
\newcommand\varphiL{{\varphi_{L}}}
\newcommand\psiL{{\psi_{L}}}
\newcommand\C{{\cos{4\psi}}}

@@

First, for an approximate approach, note that in the limit that either $t^4_\theta<<1$ or $t^4_\iota<<1$, then \eqref{eq:sigmapm} become nearly independent of that quantity except through  $\rho$ in \eqref{eq:sigmafactorrho}. Similarly if $t^4_\iota>>1$ then dependence is only through $t^4_\iota\rho$. We can approximately represent both possibilities allowing transformations which preserve $d(1+t^2_\iota)^2/(1+t^4_\iota)=2d/(\cos^2(\iota)+1)$,
\begin{align}
  \iota&\rightarrow\iota'\\
  d&\rightarrow d\frac{(1+\cos^2{\iota'})}{(1+\cos^2{\iota})}&=d.
\end{align}
Another option at the same level of approximation is preserve $d/\cos(\iota)$, but this may be less appropriately when $\cos(\iota)->0$.  In some MCMC tests, the latter form acheived better autocorrelations, although the former had slightly higher acceptance rates. A corresponding symmetry exists when $t^4_\theta<<1$.

It is also possible to explicitly formulate this symmetry in the general case, though here we are still working on a concrete plan.
To approach this, we represent subspace of the parameter space, with coordinates $\{x=\min\{t^4_\theta,t^4_\iota\},y=\max\{t^4_\theta,t^4_\iota\},C=cos(2\psi),d\}$ trivially related to our basic coordinates, but better adapted for this problem.

We will next consider an exact generalization of the distance-inclination symmetry. For this we use coordinates $\{z=xy,\Delta=(|\sigma_-|^2-|\sigma_+|^2)/\rho^2,S=|\sigma_-|^2/\rho^2,\rho\}$.  Of these latter coordinates, the last 3 will be considered fixed within the orbits of the symmetry, though only 2 combinations of these are fully specified by the observed constraints on \eqref{eq:sigmapm}. The remaining coordinate $z$ is free within $S^{1/2}-1\leq z\leq 1-\Delta$. The transformation to these coordinates is
\begin{align}
  z&=xy\\
  \Delta&=(1-x^2)(1-y^2)\\
  S&=1+x^2y^2+2xyC\\
  D&=d(1+x^{1/2})(1+y^{1/2}).
\end{align}
The symmetry is realized by any translation in $z\rightarrow z$, the new coordinates are ten recovered from the inverse transformation is 
\begin{align}
  x&=\sqrt{\Lambda(z,\Delta)-\delta(z,\Delta)}\\
  y&=\sqrt{\Lambda(z,\Delta)+\delta(z,\Delta)}\\
  C&=\frac{S-1-z^2}{2z}\\
  d&=\frac D{(1+x(z,\Delta)^{1/2})(1+y(z,\Delta)^{1/2})},
\end{align}
with $\Lambda=(1+z^2-\Delta^2)/2$ and $\delta=\sqrt{\Lambda^2-z^2}$.
Then to complete the transformation, since the argument of the complex argument of the factors in square brackets in \eqref{eq:sigmapm} will have changed, it is necessary to complete the specification by choosing $\lambda$ and $\phi$ to preserve the argument of the full expressions.

@@

We will next consider a full 2-D space of symmstry transformations, this time we use coordinates $\{{z=xy,R=(|\sigma_-/\sigma_+|^2),C,D=|\sigma_-|^{-1}}\}$.  Of these latter coordinates, $R$ and $D$ are fully specified by the observed constraints on \eqref{eq:sigmapm}, but $z$ and $C$ are free within $-1\leq C\leq1$ and $0\leq z\leq T-\sqrt{T^2-1}$ with $T=(1+C)/R-C$. The transformation to this coordinates is
\begin{align}
  z&=xy\\
  R&=\frac{x^2+y^2+2xyC}{1+x^2y^2+2xyC}\\
  C&=C\\
  D&=d\frac{(1+x^{1/2})^2(1+y^{1/2})^2}{1+x^2y^2+2xyC}.
\end{align}
The symmetry is realized by any translation in $z\rightarrow z$, the new coordinates are ten recovered from the inverse transformation is 
\begin{align}
  x&=\sqrt{\Lambda(z,\Delta)-\delta(z,\Delta)}\\
  y&=\sqrt{\Lambda(z,\Delta)+\delta(z,\Delta)}\\
  C&=C\\
  d&=D\frac {1+z^2+2zC}{(1+x(z,\Delta)^{1/2})(1+y(z,\Delta)^{1/2})},
\end{align}
with $\Lambda=(1+z^2-\Delta^2)/2$ $\Delta=(1-R)*(1+z^2+2*C)$ and $\delta=\sqrt{\Lambda^2-z^2}$.
The rest of the process follows as above.

An issue with this specification (as with the 1-D version) is that it only encompases transformations which preserve whether $t_\theta$ or $t_\iota$ is larger.  It appears that this can be remedied by starting with a parameter $q=(t_\theta/t_\iota)^{4}$ which unlike $z$ is unconstrained. The calculation is similar, using $z=T-\sqrt{T^2-1}$ with $T=((q+q^{-1})/2+C)/R-C$ as intermediates.

Since we are actually implementing this, we spell it out explicitly.  We use the following notation for convenience:
\begin{align}
  x&=t^4_{\theta}=\tan^4{\theta/2}\\
  y&=t^4_{\iota}=\tan^4{\iota/2}\\
  z&=xy\\
  \hat w&=x/y\\
  R&=|\sigma_-/\sigma_+|^2\\
  &=\frac{x^2+y^2+2xy\C}{1+x^2y^2+2xy\C}\\
  &=\frac{\hat w+\hat w^{-1}+2\C}{z+z^{-1}+2\C}\\
  \Delta&=|\sigma_+|^2-|\sigma_-|^2\\
  &=\frac{(1-x^2)(1-y^2)}{16d^2(1+x^{1/2})^2(1+y^{1/2})^2}\\
  &=\frac{(1+z^2-z(\hat w+{\hat w}^{-1})}{16d^2(1+({\hat w}z)^{1/4})^2(1+({\hat w}/z)^{1/4})^2}.
\end{align}
Using this shorthand, we transform from variables $\{\psi,x,y,d\}$ to $\{\psi,w,R,D\}$.
Of these latter coordinates, $R$ and $D$ are fully specified by the observed constraints on \eqref{eq:sigmapm},
but $\psi$ and $w$ are free. The transformation to these coordinates is
\begin{align}
  \psi&=\psi\\
  w&=\mathrm{sign}(\ln(xy))x/y\\
  R&=\frac{x^2+y^2+2xy\C}{1+x^2y^2+2xy\C}\\
  \Delta&=\frac{(1-x^2)(1-y^2)}{16d^2(1+x^{1/2})^2(1+y^{1/2})^2}.
\end{align}
Note that we have added a sign to ${\hat w}=|w|$ as a convenient trick to
store the information about whether $xy$ is greater than $1$ to provide a bijective map over the full domain. 
The symmetry is realized by any translation in $\{\psi,w\}\rightarrow\{\psi',w'\}$ preserving $R$ and $D$.
The new coordinates are ten recovered from the inverse transformation, 
\begin{align}
  \psi&=\psi\\
  x&=z\hat w\\
  y&=z/{\hat w}\\
  d&=\Delta^{-1/2}\frac {\sqrt{1+z^2-z(\hat w+{\hat w}^{-1})}}{(1+(z \hat w)^{1/4})(1+(z/{\hat w})^{1/4})},
\end{align}
with $z=T+\mathrm{sign}(w)\sqrt{T^2-1}$ and $T=(({\hat w}+{\hat w}^{-1})/(2R)+\C(1-R)/R$ as intermediates.
The Jacobian of the forward coordinate transformation its contribution to the Jacobian of the resulting
symmetry transformation are conveniently written in mixed notation,
\begin{align}
  |J|&=-2\frac{wR\Delta}{xyd}\frac{1-z^2}{1+z^2+2z\C},\\
  \frac{|J'|}{|J|}&=\frac{w'xyd}{wx'y'd'}\frac{1-z'^2}{1-z^2}\frac{1+z^2+2z\C}{1+z'^2+2z'\cos{4\psi'}}.
\end{align}

>>>>

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
\newcommand{\hatk}{\hat{k}}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand{\bsub}{\begin{subequations}}
\newcommand{\esub}{\end{subequations}}
\newcommand\betaL{{\beta_{L}}}
\newcommand\lambdaL{{\lambda_{L}}}
\newcommand\varphiL{{\varphi_{L}}}
\newcommand\psiL{{\psi_{L}}}
\newcommand\C{{\cos(4\psi)}}

@@

First, for an approximate approach, note that in the limit that either $t^4_\theta<<1$ or $t^4_\iota<<1$, then \eqref{eq:sigmapm} become nearly independent of that quantity except through  $\rho$ in \eqref{eq:sigmafactorrho}. Similarly if $t^4_\iota>>1$ then dependence is only through $t^4_\iota\rho$. We can approximately represent both possibilities allowing transformations which preserve $d(1+t^2_\iota)^2/(1+t^4_\iota)=2d/(\cos^2(\iota)+1)$,
\begin{align}
  \iota&\rightarrow\iota'\\
  d&\rightarrow d'\frac{(1+\cos^2{\iota'})}{(1+\cos^2{\iota})}.
\end{align}
Another option at the same level of approximation is preserve $d/\cos(\iota)$, but this may be less appropriately when $\cos(\iota)->0$.  In some MCMC tests, the latter form acheived better autocorrelations, although the former had slightly higher acceptance rates. A corresponding symmetry exists when $t^4_\theta<<1$.

It is also possible to explicitly formulate this symmetry in the general case.
To approach this, we represent subspace of the parameter space, with coordinates $\{x=\min\{t^4_\theta,t^4_\iota\},y=\max\{t^4_\theta,t^4_\iota\},C=\cos(4\psi),d\}$ trivially related to our basic coordinates, but better adapted for this problem.

Consider an exact generalization of the distance-inclination symmetry. For this we use coordinates $\{z=xy,\Delta=(|\sigma_-|^2-|\sigma_+|^2)/\rho^2,S=|\sigma_-|^2/\rho^2,\rho\}$.  Of these latter coordinates, the last 3 will be considered fixed within the orbits of the symmetry, though only 2 combinations of these are fully specified by the observed constraints on \eqref{eq:sigmapm}. The remaining coordinate $z$ is free within $S^{1/2}-1\leq z\leq 1-\Delta$. The transformation to these coordinates is
\begin{align}
  z&=xy\\
  \Delta&=(1-x^2)(1-y^2)\\
  S&=1+x^2y^2+2xyC\\
  D&=d(1+x^{1/2})^2(1+y^{1/2})^2.
\end{align}
The symmetry is realized by any translation in $z\rightarrow z$, the new coordinates are ten recovered from the inverse transformation is 
\begin{align}
  x&=\sqrt{\Lambda(z,\Delta)-\delta(z,\Delta)}\\
  y&=\sqrt{\Lambda(z,\Delta)+\delta(z,\Delta)}\\
  C&=\frac{S-1-z^2}{2z}\\
  d&=\frac D{(1+x(z,\Delta)^{1/2})^2(1+y(z,\Delta)^{1/2})^2},
\end{align}
with $\Lambda=(1+z^2-\Delta^2)/2$ and $\delta=\sqrt{\Lambda^2-z^2}$.
Then to complete the transformation, since the argument of the complex argument of the factors in square brackets in \eqref{eq:sigmapm} will have changed, it is necessary to complete the specification by choosing $\lambda$ and $\phi$ to preserve the argument of the full expressions.

@@

Now consider a full 2-D space of symmstry transformations,
We use the following notation for convenience:
\begin{align}
  x&=t^4_{\theta}=\tan^4{\theta/2}\\
  y&=t^4_{\iota}=\tan^4{\iota/2}\\
  z&=xy^s\\
  \hat w&=x/y^s\\
  R&=|\sigma_-/\sigma_+|^2\\
  &=\frac{x^2+y^2+2xy\C}{1+x^2y^2+2xy\C}\\
  &=\left(\frac{\hat w+\hat w^{-1}+2\C}{z+z^{-1}+2\C}\right)^s\\
  \Delta&=16(|\sigma_-|^2-|\sigma_+|^2)\\
  &=\frac{(1-x^2)(1-y^2)}{d^2(1+x^{1/2})^4(1+y^{1/2})^4}\\
  &=\frac{1-x^2-y^2+x^2y^2}{d^2(1+x^{1/2})^4(1+y^{1/2})^4}\\
  &=\frac{(xy)^{-1}-(x/y)-(x/y)^{-1}+xy}{d^2(x^{-1/4}+x^{1/4})^4(y^{-1/4}+y^{1/4})^4}\\
  &=\frac{s}{d^2}\frac{z+z^{-1}-(\hat w+{\hat w}^{-1})}{(z^{1/2}+z^{-1/2}+{\hat w}^{1/2}+{\hat w}^{-1/2})^4}
\end{align}
where $s=\pm1$.
Using this shorthand, we transform from variables $\{\psi,x,y,d\}$ to $\{\psi,w,R,D\}$.
Of these latter coordinates, $R$ and $D$ are fully specified by the observed constraints on \eqref{eq:sigmapm},
but $\psi$ and $w$ are free. The transformation to these coordinates is
\begin{align}
  \psi&=\psi\\
  w&=\mathrm{sign}(\ln(xy))x/y^s\\
  R&=\frac{x^2+y^2+2xy\C}{1+x^2y^2+2xy\C}\\
  \Delta&=\frac{(1-x^2)(1-y^2)}{d^2(1+x^{1/2})^4(1+y^{1/2})^4}.
\end{align}
Note that we have added a sign to ${\hat w}=|w|$ as a convenient trick to
store the information about whether $xy$ is greater than $1$ to provide a bijective map over the full domain. 
Also, swapping the sign $s$, which we have not yet specified, reverses the roles of $w$ and the intermediate quantitity $z$.
The symmetry is realized by any translation in $\{\psi,w\}\rightarrow\{\psi',w'\}$ preserving $R$ and $D$.
The new coordinates are ten recovered from the inverse transformation, 
\begin{align}
  \psi&=\psi\\
  x&=\sqrt{z\hat w}\\
  y&=\sqrt{z/{\hat w}}\\
  d&={\Delta^{-1/2}}\frac {\sqrt{1+z^2-z(\hat w+{\hat w}^{-1})}}{(z^{1/2}+z^{-1/2}+{\hat w}^{1/2}+{\hat w}^{-1/2})^2}
\end{align}
with $z=T+\mathrm{sign}(w)\sqrt{T^2-1}$ and $T=(({\hat w}+{\hat w}^{-1})/(2R^s)+\C(1-R^s)/R^s$ as intermediates.
Choosing $s=\mathrm{sign}(\ln(R))$, ensures that $0\leq R^s\leq1$, so that $z$ remains real. 
The Jacobian of the forward coordinate transformation its contribution to the Jacobian of the resulting
symmetry transformation are conveniently written in mixed notation,
\begin{align}
  |J|&=4\frac{wR\Delta}{xyd}\frac{z-z^{-1}}{z+z^{-1}+2\C},\\
  \frac{|J|}{|J'|}&=\frac{wx'y'd'}{w'xyd}\frac{z-z^{-1}}{z'-z'^{-1}}\frac{z'+z'^{-1}+2\cos(4\psi')}{z+z^{-1}+2\cos(4\psi)}.
\end{align}
